% Pandoc LaTeX template for course handouts
% Usage: pandoc input.md -o output.tex --template=handout_template.tex --listings
\documentclass[10pt,a4paper]{article}

% Geometry
\usepackage[margin=2.5cm]{geometry}

% Fonts and encoding
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}

% Math
\usepackage{amsmath,amssymb,amsfonts}

% Tables
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}

% Graphics
\usepackage{graphicx}

% Colors
\usepackage{xcolor}
\definecolor{mlpurple}{RGB}{51,51,178}
\definecolor{mlblue}{RGB}{0,102,204}
\definecolor{codebg}{RGB}{248,248,248}
\definecolor{codeframe}{RGB}{200,200,200}

% Code listings
\usepackage{listings}
\lstset{
  backgroundcolor=\color{codebg},
  basicstyle=\ttfamily\small,
  breakatwhitespace=false,
  breaklines=true,
  captionpos=b,
  commentstyle=\color{gray},
  frame=single,
  framerule=0.5pt,
  rulecolor=\color{codeframe},
  keepspaces=true,
  keywordstyle=\color{mlblue}\bfseries,
  language=Python,
  numbers=none,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  stringstyle=\color{mlpurple},
  tabsize=4,
  xleftmargin=0.5em,
  xrightmargin=0.5em,
  aboveskip=1em,
  belowskip=1em
}

% Hyperlinks
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  linkcolor=mlblue,
  urlcolor=mlblue,
  citecolor=mlblue
}

% Headers and footers
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhf{}
\lhead{\textcolor{mlpurple}{\textbf{Responsible AI}}}
\rhead{\textcolor{gray}{Basic}}
\cfoot{\thepage}
\renewcommand{\headrulewidth}{0.4pt}
\renewcommand{\footrulewidth}{0pt}

% Paragraph spacing
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.5em}

% Section formatting
\usepackage{titlesec}
\titleformat{\section}
  {\Large\bfseries\color{mlpurple}}
  {\thesection}{1em}{}
\titleformat{\subsection}
  {\large\bfseries\color{mlblue}}
  {\thesubsection}{1em}{}
\titleformat{\subsubsection}
  {\normalsize\bfseries}
  {\thesubsubsection}{1em}{}

% Lists
\usepackage{enumitem}
\setlist[itemize]{topsep=0.5em, itemsep=0.25em}
\setlist[enumerate]{topsep=0.5em, itemsep=0.25em}

% Tight list support for pandoc
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

% Title
\title{\Huge\bfseries\color{mlpurple} Handout 1: Basic AI Ethics Checklist}
\author{Machine Learning for Smarter Innovation}
\date{}

\begin{document}

\maketitle
\thispagestyle{fancy}

\section{Handout 1: Basic AI Ethics
Checklist}\label{handout-1-basic-ai-ethics-checklist}

\subsection{For Beginners}\label{for-beginners}

\subsubsection{Before Building an AI
System}\label{before-building-an-ai-system}

\textbf{Purpose Check:} - {[} {]} Is there a real need for this AI
system? - {[} {]} Have we considered non-AI alternatives? - {[} {]} Can
we explain the value to stakeholders?

\textbf{People Check:} - {[} {]} Who will be affected by this system? -
{[} {]} Have we talked to users and affected communities? - {[} {]} Who
might be harmed?

\textbf{Data Check:} - {[} {]} Do we know where our data came from? -
{[} {]} Do we have permission to use this data? - {[} {]} Is our data
representative of all users? - {[} {]} Have we checked for obvious
biases in the data?

\subsubsection{During Development}\label{during-development}

\textbf{Fairness Check:} - {[} {]} Have we tested performance across
different groups? - {[} {]} Are error rates similar for all groups? -
{[} {]} Have we documented any performance differences?

\textbf{Transparency Check:} - {[} {]} Can we explain how the system
makes decisions? - {[} {]} Have we documented what data we use? - {[}
{]} Is it clear to users that AI is being used?

\textbf{Safety Check:} - {[} {]} What happens if the system makes a
mistake? - {[} {]} Have we tested edge cases? - {[} {]} Is there a human
review process?

\subsubsection{Before Deployment}\label{before-deployment}

\textbf{Review Check:} - {[} {]} Has someone outside the team reviewed
the system? - {[} {]} Have we tested with real users? - {[} {]} Do we
have a plan for monitoring after launch?

\textbf{Documentation Check:} - {[} {]} Have we created a model card? -
{[} {]} Are limitations clearly documented? - {[} {]} Do users know how
to report problems?

\textbf{Legal Check:} - {[} {]} Have we checked relevant regulations
(GDPR, AI Act)? - {[} {]} Do we have proper consent from data subjects?
- {[} {]} Have we consulted legal counsel?

\subsubsection{After Deployment}\label{after-deployment}

\textbf{Monitoring Check:} - {[} {]} Are we tracking system performance
regularly? - {[} {]} Have we set up alerts for problems? - {[} {]} Are
we collecting user feedback?

\textbf{Response Check:} - {[} {]} Do we have a process for handling
complaints? - {[} {]} Can users appeal AI decisions? - {[} {]} Are we
fixing identified problems quickly?

\subsubsection{Red Flags (STOP if any of these are
true)}\label{red-flags-stop-if-any-of-these-are-true}

\begin{itemize}
\tightlist
\item
  \textbf{No} - We don't know who provided our training data
\item
  \textbf{No} - We haven't tested performance across different
  demographic groups
\item
  \textbf{No} - We can't explain how the system makes decisions
\item
  \textbf{No} - There's no human oversight for high-stakes decisions
\item
  \textbf{No} - We have no plan for monitoring after deployment
\item
  \textbf{No} - We don't have user consent for data usage
\item
  \textbf{No} - The system makes decisions that significantly affect
  people's lives without review
\end{itemize}

\subsubsection{Quick Terminology}\label{quick-terminology}

\begin{itemize}
\tightlist
\item
  \textbf{Bias}: Systematic errors that favor certain groups over others
\item
  \textbf{Fairness}: Treating similar people similarly
\item
  \textbf{Transparency}: Being open about how AI systems work
\item
  \textbf{Accountability}: Clear responsibility for AI outcomes
\item
  \textbf{Model Card}: A document describing an AI model's capabilities
  and limitations
\end{itemize}

\subsubsection{Resources for Beginners}\label{resources-for-beginners}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Google's People + AI Guidebook}: Free, visual, practical
\item
  \textbf{IBM AI Ethics Toolkit}: Simple templates and checklists
\item
  \textbf{Microsoft Responsible AI Resources}: Video tutorials and
  guides
\item
  \textbf{EU AI Act Summary} (official website): Plain language version
\end{enumerate}

\subsubsection{Key Takeaway}\label{key-takeaway}

\textbf{Ethics isn't optional - it's essential for building AI systems
that work for everyone.}

Start with this checklist for every AI project, no matter how small.

\end{document}
